Ensemble Stability of Biological Networks

### 2014-09-15 paper.tex
introduce parameters and work on char poly section

@@

\section{System graph and decomposition into strongly connected components}
We will show how the stability of the dynamical system associated to Jacobian $A$ is characterized by properties of the associated graph $G_A$. The characteristic polynomial $\chi(A)$ associated to $A$ can be factored according to the strongly connected components of the graph $G$. $$\frac{\chi(M)'(z)}{\chi(M)(z)} = \tr (zI - M)^{-1}$$ where, $\chi(M)(z) = \det (zI - M)$ is the chararacteristic polynomial of the matrix $M$ with $z$ a complex variable.  This formula is readily verified when $M$ is diagonal and has the nice property of being additive over decompositions of the matrix. Making the change of variables $x=\frac{1}{z}$, we have $$\tr (I - xM)^n = \sum_{n=0}^\infty x^n \sum_{a \in \Vertex(G)} \sum_{p \in \Path_n (G)(a,a)} F(p)$$
where $G$ is the graph of the matrix and $\Path_n (G;a,b)$ is the set of
paths of length $n$ from $a$ to $b$.  $(\Vertex(G), \Path(G), \cdot)$ is a category; $F$ is the functor from this category to the multiplicative semigroup of the real numbers which corresponds to the matrix $M$.

@@

Assume that $M$ is stable and that it has $2m$ complex roots which come in complex conjugate pairs and $n-m$ real roots.  Let $r_1 \ldots, r_{2m}$ be the complex roots with $r_{2k+1}$ being the complex conjugate of $r_{2k}$ and let $r_{2m+1}, \ldots,  r_{n}$ be the real roots. Then we can write the factorization as follows:
\[
\chi(M)(z) = \prod_{k=1}^{m} (z^2 - (r_{2k} + r_{2k+1}) z + r_{2k} r_{2k+1})
             \prod_{k=2m+1}^{n} (z - r_k)
\]
Since $r_{2k}$ and $r_{2k+1}$ are complex conjugates, $r_{2k} + r_{2k+1} =
\Re r_{2k} = \Re r_{2k+1}$ and $r_{k} r_{2k+1} > 0$.   Since all the
roots have negative real part, all the coefficients of the terms in
the products above are positive so, when we multiply out the product,
we will have all positive terms.  Hence, we conclude that a necessary
condition for $M$ to be stable is that $s_k > 0$ for all $k$.

@@

@@

\section{Appendix A}\label{sec:appA}
Suppose that we have a family
of polynomials whose roots are of one type but turn into another type.
There are three ways this could happen:
\begin{itemize}
\item{1.} A real root could go from being positive to being negative
  or vice versa.  For this to happen, we must pass through an
  intermediate value with a zero root.  For that intermediate value,
  we have $D = 0$.
\item{2.} A pair of real roots could turn into a pair of complex roots
  or vice versa.  For this to happen, we must pass through an
  intermediate value with a repeated root.  For that intermediate
  value, we have $\delta = 0$.
\item{3.} A conjugate pair of complex roots could go from having
  positive real part to having negative real part or vice versa.  For
  this to happen, we must pass through an intermediate value in which
  the real part is zero.  For that intermediate value, the two roots
  are imaginary so, bing conjugates, they sum to zero, hence we have
  $\Delta = 0$.
\end{itemize}
\noindent  Thus, we conclude that the boundaries of our regions
$R_{abcd}$ consist of portions of the surfaces $D=0, \delta=0,
\Delta=0$ and we will characterize them by studying these surfaces.

>>>>

In order to accommodate the possibility of uncertainty in our modelling,
we will generalize our notion of a system on a network to that of a
system with random parameters.  Again, this will involve three
steps. First, we provide a measure space $S$ which represents the
values over which our parameters can vary. Next, instead of a single
vector field, we consider a family of vector fields parameterized by
this space. That is to say, we introduce a mapping $V$ from $S$ to
sections of the tangent bundle of the product of manifolds. As before,
for each point $p \in S$, the vector field $V(p)$ must be consistent with
the system graph. Finally, we select a probability measure $\mu$ on
$S$ which represents our understanding of which values of the parameters
are likely. In accord with Bayesian statistics, we might revise this
distribution as data comes in or use it to estimate parameters and error
bars from experimental results.

Having done this, we are now in a position to do a probabilistic analysis.
Given some quantity $q$ characterizing our system (such as, for instance,
the time it would take a particle to travel between two points), this
quantity now becomes a random variable.  We may then study the average
value, variance, and the like.  Another question one can ask is that of
robustness, which is the conditional probability that, if some property
holds for a set of parameters, it will still hold if we make some random
perturbation about those parameter values.

@@

@@

\begin{lemma}
If $M$ is stable, then $s_i > 0$ for all $i$ between $1$ and $n$.
\end{lemma}

\begin{proof}
Assume that $M$ is stable and that it has $2m$ complex roots which come in
complex conjugate pairs and $n-m$ real roots.  Let $r_1 \ldots, r_{2m}$ be
the complex roots with $r_{2k+1}$ being the complex conjugate of $r_{2k}$
and let $r_{2m+1}, \ldots,  r_{n}$ be the real roots. Then we can write the
factorization as follows:
\[
\chi(M)(z) = \prod_{k=1}^{m} (z^2 - (r_{2k} + r_{2k+1}) z + r_{2k} r_{2k+1})
             \prod_{k=2m+1}^{n} (z - r_k)
\]
Since $r_{2k}$ and $r_{2k+1}$ are complex conjugates, $r_{2k} + r_{2k+1} =
\Re r_{2k} = \Re r_{2k+1}$ and $r_{k} r_{2k+1} > 0$.   Since all the
roots have negative real part, all the coefficients of the terms in
the products above are positive so, when we multiply out the product,
we will have all positive terms.  Hence, we conclude that a necessary
condition for $M$ to be stable is that $s_k > 0$ for all $k$.
\end{proof}

To state the next criterion, we will require the quantity $\Delta$, which
is defined as $\prod_{j=1}^{n-1} \prod_{k=j+1}^n (r_1 + r_j)$.  Since
this expression is invariant under permuting the roots, it may be expressed
as a function of the coefficients $s_i$; for small values of $n$, the
resulting expression looks as follows:
\begin{align*}
n=2 \qquad &\Delta = s_1 \\
n=3 \qquad &\Delta = s_3 - s_1 s_2 \\
n=4 \qquad &\Delta = s_1 s_2 s_3 - s_3^2 - s_1^2 s_4 \\
n=5 \qquad &\Delta = s_1 s_2 s_3 s_4 - s_3^2 s_4 - s_1^2 s_4^2 -
                     s_1 s_2^2 s_5 + s_2 s_3 s_5 + 2 s_1 s_4 s_5 - s_5^2
\end{align*}

\begin{lemma}
Given a polynomial $p(z) = z^n + \sum_{k=1}^{n-1} s_k z^{n-k}$ and a real number
$\epsilon > 0$, there exists $\delta \in \mathbb{R}$ such that, if $p'(x) =
x^n + \sum_{k=1}^{n-1} {s'}_i z^i$ is a polynomial for which $|s_i - {s'}_i|
< \delta$ then to every root $r$ of $p$ there corresponds a root $r'$ of
$p'$ such that $|r-r'| < \epsilon$.
\end{lemma}

\begin{proof}
Without loss of generality, we may assume that $\epsilon < 1$.

We first consider the case where $p = x^n$.

If we make the rescaling $x = \epsilon y/2$ in the equation $p(x)  = 0$,
we obtain the equation
\[
y^n + \sum_{k=1}^{n-1} \left(\frac{2}{\epsilon}\right)^k s_k y^{n-k} = 0 .
\]
Cauchy's bound on the roots of this equation is
\[
|y| < 1 + \max_j \left(\frac{2}{\epsilon}\right)^j |s_j| ,
\]
which becomes the bound
\[
|x| < \frac{\epsilon}{2} + \max_j \left(\frac{2}{\epsilon}\right)^{j-1} |s_j|
< \frac{\epsilon}{2} + \left(\frac{2}{\epsilon}\right)^{n-1} \max_j |s_j|
\]
for the original equation.  Hence, if we choose $\delta = (\epsilon/2)^{n}$,
we will have $|x| < \epsilon$.

Next, we generalize to the case $p = (x-r)^n$.

To extend this result of the general case, we note that, given two
polynomials $p = x^m + \sum_{k=1}^n a_k x^{m-k}$ and
$q = x^n + \sum_{k=1}^n b_k x^{n-k}$, the Jacobian of the map which
sends the coefficients of the matrices to the coefficients of the
product $pq$ is the Sylvester matrix, whose determinant is the resultant
of $p$ and $q$.  Hence, is $p$ and $q$ have no common root, the map
is locally invertible.
\end{proof}

\begin{lemma}
If a point lies on the boundary of the set of stable polynomials, then
either $s_n = 0$ and $\Delta = 0$.
\end{lemma}

\begin{proof}
Since the mapping which sends the roots of a polynomial to its
coefficients has non-singular Jacobian when the discriminant is not
zero, if a stable polynomial has distinct roots, it does not lie on
the boundary.

Suppose that we have a family
of polynomials whose roots are of one type but turn into another type.
There are three ways this could happen:
\begin{itemize}
\item{1.} A real root could go from being positive to being negative
  or vice versa.  For this to happen, we must pass through an
  intermediate value with a zero root.  For that intermediate value,
  we have $D = 0$.
\item{2.} A pair of real roots could turn into a pair of complex roots
  or vice versa.  For this to happen, we must pass through an
  intermediate value with a repeated root.  For that intermediate
  value, we have $\delta = 0$.
\item{3.} A conjugate pair of complex roots could go from having
  positive real part to having negative real part or vice versa.  For
  this to happen, we must pass through an intermediate value in which
  the real part is zero.  For that intermediate value, the two roots
  are imaginary so, bing conjugates, they sum to zero, hence we have
  $\Delta = 0$.
\end{itemize}
\noindent  Thus, we conclude that the boundaries of our regions
$R_{abcd}$ consist of portions of the surfaces $D=0, \delta=0,
\Delta=0$ and we will characterize them by studying these surfaces.

To complete the proof, we show that the condition $\delta = 0$ is not
needed to describe the boundary.  Suppose that $\delta = 0$.  Then there
must exist a repeated root.  If the real part of any root, repeated or
not, is positive, then the polynomial is not stable and, by continuity,
if one allows the coefficients to vary within a sufficiently small
ball about that solution, the root will remain negative, so the
polynomial cannot lie on the boundary of the stable region.  If all
the roots, including the repeated ones, are negative then, by a similar
argument the point is in the interior of the stable region.  The only
other possibility is that some root has zero real part, but then
$\Delta = 0$ so the condition $\delta = 0$ is superfluous.
\end{proof}

@@

\section{System graph and decomposition into strongly connected components}
Next, we will want to relate prpoperties of our matrix and its eigenvalues
to  properties of the associated graph $G_A$.  To do this, we will first
examine the category of paths on $G_A$, and derive a formula expressing
the chararacteristic polynomial as a sum over cycles.

Given a digraph $G$, let $\Path (a,b)$ denote the set of paths from the
vertex $a$ to the vertex $b$.  Given a path $p_1$ from $a$ to $b$ and a path
$p_2$ from $b$ to $c$, we may concatenate them to obtain a path $p_1 \circ
p_2$ from $a$ to $c$.  Hence, $(\Vertex(G), \Path(G), \circ)$ forms a category.

The characteristic polynomial $\chi(A)$ associated to $A$ can be factored according to the strongly connected components of the graph $G$. $$\frac{\chi(M)'(z)}{\chi(M)(z)} = \tr (zI - M)^{-1}$$ where, $\chi(M)(z) = \det (zI - M)$ is the chararacteristic polynomial of the matrix $M$ with $z$ a complex variable.  This formula is readily verified when $M$ is diagonal and has the nice property of being additive over decompositions of the matrix. Making the change of variables $x=\frac{1}{z}$, we have $$\tr (I - xM)^n = \sum_{n=0}^\infty x^n \sum_{a \in \Vertex(G)} \sum_{p \in \Path_n (G)(a,a)} F(p)$$
where $G$ is the graph of the matrix and $\Path_n (G;a,b)$ is the set of
paths of length $n$ from $a$ to $b$.  ; $F$ is the functor from this category to the multiplicative semigroup of the real numbers which corresponds to the matrix $M$.

@@

\section{Appendix A}\label{sec:appA}
